Variance reduction for MCMC methods via martingale representations

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

@@

Let us sketch the main steps of the proof. First using integration by parts we prove that

@@

A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]

@@

stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive

>>>>

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\label{eq:expression-variance}
\PVar_x[\pi_{K,n}^N(f)]=\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Estimation of coefficients}
The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
(the so-called ``training paths''):
\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
for \(r=1,\ldots, n-1,\)
where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
Next  we estimate the coefficients \( \bar a_{l,k}\) using
the formula
\label{eq:a_est_int}
\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories. In many cases integration can be done in closed analytical form. For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\widehat M_{K,n}^N(f) =\frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
\leq \sum_{k=1}^{K}\frac{1}{n^2}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

@@

Let us sketch the main steps of the proof. First using integration by parts, we prove that

@@

A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],

@@

stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive
